Neural Circuit Organization of the Ctenophore Apical Organ

### 2025-03-23 manuscript/Manuscript.qmd
MS: edit balancer imaging section

Our synapse annotation revealed synapses between the ANN neurons and synapses that ANN_Q1-4 forms on itsels (autapses). The ANN neurons also form synapses on the gravity-sensing balancer cells.

@@

From ANN_Q1Q2, synapses were formed with 6 of the 37 balancer cells in the Q1 region, and 8 of the 32 balancer cells in the Q2 region. Similarly, from ANN_Q3Q4, synapses were formed with 1 of the 32 balancer cells in the Q3 region, and 5 of the 28 balancer cells in the Q4 region. ANN_Q1-4 formed input synapses with balancer cells in the Q1 region (7 cells), Q2 region (11 cells), Q3 region (6 cells), and Q4 region (10 cells). Some balancer cells received inputs from both ANN_Q1Q2 or ANN_Q3Q4 and ANN_Q1-4. While previous studies have suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez-Nicaise1974], our data did not reveal any synaptic inputs from balancer cells to ANN neurons. However, we identified "bridge cells" that actively formed input synapses with ANN neurons, highlighting their potential role in the network.

### Identification of Neuron-like Bridge Cells Forming Afferent Synapses in the Gravity-Sensitive Neural Circuit

From our data, we identified several cell groups that form multiple afferent synapses with ANNs. These cells, based on their morphological features, were found to be the bridge cells first described by Tamm et al. in 2002 [@Tamm_2002]. These cells are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge. They originate from the base of paired balancer cells along the tentacle surface and extend across the sagittal plane toward the base of the opposite balancer cells. In regions where the mitochondria of the bridge cells are localized (approximately 30%), a presynaptic triad structure, similar to that of ANNs, was found, containing synaptic vesicles and smooth endoplasmic reticulum.

Our three-dimensional reconstruction data revealed that the bridge cells form two distinct cell groups across the sagittal plane, between the Q1Q2 and Q3Q4 regions. In the Q1Q2 region, 14 cells were identified, while in the Q3Q4 region, 12 cells were identified, totaling 26 bridge cells. Nearly all of these bridge cells (25 out of 26) exhibited afferent synapses from ANNs. For bridge cells located in the Q1Q2 region, synaptic inputs came primarily from ANN_Q1Q2 (11 cells), ANN_Q1-4 (1 cell), or both (2 cells). For bridge cells located in the Q3Q4 region, synaptic inputs were received from ANN_Q3Q4 (1 cell), ANN_Q1-4 (7 cells), or both (1 cell). In other words, bridge cells in the Q1Q2 region mainly received input from ANN_Q1Q2, while those in the Q3Q4 region received input primarily from ANN_Q1-4, showing a distinct pattern of input in the two regions.

Some bridge cells also formed afferent synapses with ANN neurons. For example, bridge cells in the Q1Q2 region (3 cells) formed synapses with ANN_Q1Q2 or (2 cells) both ANN_Q1Q2 and ANN_Q1-4. Bridge cells in the Q3Q4 region formed synapses with ANN_Q3Q4 (1 cell) or ANN_Q1-4 (6 cells). A notable difference was observed between the Q1Q2 and Q3Q4 regions in the proportion of synaptic inputs from bridge cells to ANN neurons.

Interestingly, in both regions, bridge cells also formed synapses with adjacent bridge cells. However, no synaptic input was found from bridge cells across the sagittal plane to those in the opposite region. To analyze the grouped synaptic connectivity graph, we classified cell types within each region, collapsed cells of the same type into a single node, summed the number of synapses, and explored directed pathways to effectors (balancer cell groups). The results revealed a feedback pathway from ANN neurons through the synaptic connections formed by bridge cells, thereby shedding light on the neural circuit structure involving these cells.

@@

### Control of Balancer Ciliary Movements by Gravity-Sensitive Neural Circuits

To investigate the function of gravity-sensitive circuits, we conducted a comparative analysis of balancer ciliary movements across different regions. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Furthermore, it has been suggested that differences in the asymmetry of statolith morphology and the elongation direction of balancer cilia between the tentacular and sagittal planes could result in variations in the force received from the statolith [@Tamm_2014; @Tamm2015]. Building on these findings, we standardized conditions for observing balancer cilia by tilting the microscope by 90 degrees and fixing samples such that the aboral-oral axis was aligned parallel to the vertical stage, with the aboral side facing upward. Using connectome analysis, we aimed to evaluate the neural circuit differences by analyzing the control patterns of left and right balancer cilia in the tentacular and sagittal planes. Specifically, we hypothesized that differences in control between Q1 (Q2) and Q3 (Q4) regions, as well as between Q1 (Q4) and Q2 (Q3) regions, would reveal distinct functions of ANNs (e.g., ANN Q1Q2/Q3Q4 vs. ANN Q1-4). For the sagittal plane, we recorded balancer ciliary movements in the Q1(4) and Q2(3) regions (left and right sides of the image) and, for the tentacular plane, in the Q1(3) and Q2(4) regions, using a high-speed camera (100 fps) for 2 minutes (12,000 frames). Regions of interest (ROIs) were defined in areas where brightness changes indicated ciliary beating. We measured these brightness changes to generate dynamics graphs of balancer ciliary movements for both planes.

During the 2-minute recordings, balancer cilia exhibited movements that could be fast, slow, or stop abruptly (arrest) and start moving again (re-beat). Occasionally, large contraction responses caused the entire cydippid to move out of frame, and data from these moments were excluded. We focused on two metrics for left-right comparison: (1) the timing of ciliary arrest and (2) the timing of re-beat initiation following an arrest. Due to the low baseline beating frequency of the cilia, rapid synaptic input responses were not evaluated, suggesting that ciliary beat regulation may involve non-synaptic control mechanisms. The analysis revealed that both arrest and re-beat events were synchronized between left and right balancers in both planes within a few milliseconds, suggesting that these events are governed by synaptic input from neurons. Interestingly, while the re-beat timing showed no significant differences between the sagittal and tentacular planes, arrest timing exhibited greater variability in the tentacular plane. Comparing these findings with neural circuit diagrams suggests that in the sagittal plane, the same ANNs (Q1Q2 or Q3Q4) project to both balancer cell groups, transmitting signals simultaneously and enabling synchronization. In contrast, in the tentacular plane, separate ANNs project to different regions, potentially introducing an additional synaptic step, which may account for the observed variability in arrest timing.

>>>>

Our synapse annotation revealed synapses between the ANN neurons and synapses that ANN_Q1-4 forms on itself (autapses). The ANN neurons also form synapses on the gravity-sensing balancer cells.

@@

ANN_Q1-4 formed synapses on balancer cells in all four quadrants (on 7 cells in Q1, 11 cells in Q2, 6 cells in Q3 and 10 cells in Q4) while ANN_Q1Q2 and ANN_Q3Q4 synapsed to balancer cells in their respective quadrants (ANN_Q1Q2 on 6/37 cells in Q1 and 8/32 cells in Q2; ANN_Q3Q4 on 1/32 cells in Q3 and 5/28 cells in Q4). Some balancer cells received inputs from both ANN_Q1-4 and either of ANN_Q1Q2 or ANN_Q3Q4. While previous studies suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez-Nicaise1974], our data did not reveal any presynaptic sites in balancer cells.

The second group of cells that form synaptic contact with the ANN were the bridge cells. Bridge cells, first described by Tamm et al. in 2002 [@Tamm_2002], are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge. Their somata are located at the base of the paired balancer-cell clusters along the tentacle surface and extend across the sagittal plane towards the base of the opposite balancer cells. Bridge cells form two distinct cell groups across the sagittal plane, in the Q1Q2 and Q3Q4 regions. In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12 cells.

Bridge cells had presynaptic sites with the typical presynaptic triad structure near 30% of their mitochondria. These bridge-cell synapses were formed on ANN neurons or other bridge cells. Bridge cells in the Q1Q2 region formed synapses on ANN_Q1Q2 (3 cells) or on both ANN_Q1Q2 and ANN_Q1-4 (2 cells). Bridge cells in Q3Q4 synapsed on ANN_Q3Q4 (1 cell) or ANN_Q1-4 (6 cells).

Nearly all bridge cells (25/26) also received synaptic input from ANNs. Bridge cells in Q1Q2 received inputs from ANN_Q1Q2 (11 cells), ANN_Q1-4 (1 cell), or both (2 cells). In Q3Q4, synapses on bridge cells were from ANN_Q3Q4 (1 cell), ANN_Q1-4 (7 cells), or both (1 cell).

In both The Q1Q2 and the Q3Q4 regions, bridge cells also formed synapses with adjacent bridge cells. However, we found no synapses across the sagittal plane to bridge cells in the opposite region.

To analyze the synaptic connectivity graph of the balancer organ and the ANN, we grouped cells of the same type and in the same region into a single node, summed the number of synapses, and laid out the network to reflect the anatomy of the four quadrants (Figure 3H). The network is characterised by feedback connections among ANN neurons and between ANN and bridge cells with a clear rotational (or mirror) symmetry. Q1Q2 and Q3Q4 also form separate local circuits centered around their regional ANN neurons while ANN_Q1-4 mediates global connectivity across the entire balancer organ. Notably, there were no synapses from the mechanosensory balancer cells to the ANN or from ANN to motor cells (e.g. ciliary groove), contrary to our initial expectation of a sensory-motor or input-output model of balancer function.

The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, EF cells and several non-ciliated, monociliated or biciliated cells. These cell types and synaptic connections are outside the gravisensory organ and will be described elsewhere.

@@

### Dynamics of balancer cilia imply a coordinating function for the nerve net

To investigate the regulation of balancer function, we carried out a high-speed video-microscope analysis of balancer cilia. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014; @Tamm2015]. To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector. We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage.

In larvae with their sagittal plane facing the objective, we could compared balancer-cilia movements between Q1(4) and Q2(3). In other larvae oriented in the tentacular plane, we could simultaneously image Q1(3) and Q2(4). We used a high-speed camera and recorded at 100 fps for 2 minutes (12,000 frames). To analyse ciliary beating, we selected regions of interest (ROIs) in areas where brightness changes indicated ciliary beating. Ciliary beating was both manually quantified and plotted as kymographs.

During the two-minute recordings, balancer cilia could beat fast, slow, or exhibit abrupt stops (arrest) and start moving again (re-beat). Occasionally, large body-contractions moved the entire cydippid out of frame, and data from these episods were excluded. We focused on three metrics for inter-quadrant comparison: (1) the timing of ciliary arrest, (2) the timing of re-beat after an arrested phase, and (3) ciliary beat frequency.

We found that arrest and re-beat events were synchronized between balancers across the sagittal plane. However, while the timing of re-beat was also near-simultaneous along the tentacular plane, arrest timing was offset by up to 2.xy <!-- add max value--> seconds (Figure 4) <!-- save source data for all plots in manuscript/source_data with the naming convention e.g. Figure 1A--source data 1 -->. Overall, these data reveal that arrests are only coordinated between Q1-Q2 and Q3-Q4, whereas re-beat is coordinated over all four quadrants.

Evaluating these data in light of the circuit diagram suggests that shared ANN inputs (ANN_Q1Q2 or ANN_Q3Q4) to a pair of balancers along the sagittal plane may underlie their synchronized arrests. In contrast, in the tentacular plane, separate ANNs innervate the balancer pairs. At the same time, the ANN_Q1-4 neuron synapses on all four balancers, hinting at a neural substrate for their synchronised re-beat.
